Future Directions

### 2021-02-23 manuscripts/laplacian_smoothing/AISTATS/minimax_optimal_laplacian_smoothing.tex
Detailed pass over the main AISTATS paper

\documentclass[twoside]{article}
\usepackage[accepted]{aistats2021}
\usepackage{amsmath, amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{verbatim}

@@

\newcommand{\Linv}{L^{\dagger}}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}

@@

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\theoremstyle{remark}

@@

	\aistatsauthor{Alden Green, Sivaraman Balakrishnan, and Ryan J. Tibshirani}

	\aistatsaddress{Carnegie Mellon University}]

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to non-parametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator $\wh{f}$, and a goodness-of-fit test based on $\wh{f}$. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, where $\Xset \subset \Rd$ and $1 \leq d < 4$; in the estimation problem, they are within a $\log n$ factor of optimal when $d = 4$. Additionally, we prove that Laplacian smoothing is manifold adaptive: if $\Xset$ is an $m$ dimensional manifold embedded in $\Reals^d$ with $m \ll d$, then the error rate of Laplacian smoothing methods depends only on the intrinsic dimension $m$ and not on the ambient dimension $d$. 
\end{abstract}

\section{INTRODUCTION}
In the nonparametric regression problem, we observe samples $(X_1,Y_1),\ldots,(X_n,Y_n)$ according to the signal plus noise model
\begin{equation}
\label{eqn:signal_plus_noise_model}
Y_i = f_{0}(X_i) + \varepsilon_i,
\end{equation}
with $\varepsilon_i \sim N(0,1)$ independent Gaussian noise. Our goal is to perform statistical inference on the unknown regression function $f_0$, by which we mean either \emph{estimating} $f_0$ or more simply \emph{testing} whether $f_0 = 0$, i.e whether there is any signal present.

The Laplacian smoothing estimator $\wh{f}$ \citep{smola2003} is a penalized least squares estimator, defined on a graph. Letting $G = \bigl(\{1,\ldots,n\},W\bigr)$ be a weighted graph over vertices $\{1,\ldots,n\}$ which we associate with the sample points $\{X_1,\ldots,X_n\}$, the estimator $\wh{f}$ is given by
\begin{equation}
\label{eqn:laplacian_smoothing}
\wh{f} :=  \min_{f \in \Reals^n} \biggl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T \Lap f \biggr\}.
\end{equation}
Here $\Lap$ is the graph Laplacian matrix (defined in Section~\ref{sec:problem_setup_and_background}), $G$ is typically a geometric graph (such as a $k$NN or neighborhood graph) and the penalty
\begin{equation*}
f^T \Lap f = \frac{1}{2} \sum_{i,j = 1}^{n} W_{ij}\bigl(f_i - f_j\bigr)^2
\end{equation*}
thus encourages $\wh{f}_i \approx \wh{f}_j$ when $X_i \approx X_j$. Assuming~\eqref{eqn:laplacian_smoothing} is a reasonable estimator of $f_0$, the statistic
\begin{equation}
\label{eqn:laplacian_smoothing_test}
\wh{T} := \frac{1}{n}\bigl\|\wh{f}\bigr\|_2^2 
\end{equation}
is in turn a reasonable way to decide if $f_0 = 0$.

Of course there exist many methods for nonparametric regression \citep{gyorfi2006,wasserman2006,tsybakov2008_book}, but Laplacian smoothing has its own set of advantages. For instance:
\begin{itemize}
	\item \emph{Computational ease.} Laplacian smoothing is fast, easy, and stable to compute. The estimate $\wh{f}$ can be computed by solving a symmetric, diagonally dominant linear system. There are known extremely fast and reliable solvers for exactly this problem (see e.g. the seminal papers of \citet{spielman2011,spielman2013,spielman2014}, or the monograph by~\citet{vishnoi2012} and references therein).
	\item \emph{Generality.} Laplacian smoothing is well defined whenever one can associate a graph with some observed responses. This generality lends itself to many different data modalities: for instance text and image classification, to name only two \citep{kondor2002, belkin03a,belkin2006}. 
	\item \emph{Weak supervision.} Although we study Laplacian smoothing in the supervised learning setup~\eqref{eqn:signal_plus_noise_model}, it can be easily adapted to the semi-supervised or unsupervised settings \citep{nadler09,slepcev17,calder2019b,dunlop2020}.
\end{itemize}

For these reasons, a body of work has emerged analyzing the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, these works can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item In the \emph{fixed design perspective}, the design points $\{X_1,\ldots,X_n\}$ and the graph $G$ are treated as fixed, and inference is carried out on the evaluations $\bigl(f_0(X_1),\ldots,f_0(X_n)\bigr) \in \Reals^n$.\footnote{To be clear, while in this setting $G$ may be a geometric graph, in principle it could be any graph over $\{1,\ldots,n\}$.} In this context, tight upper bounds have been derived on the error of graph-based estimators \citep{wang2016, sadhanala16,sadhanala17,kirichenko2017,kirichenko2018} and tests \citep{sharpnack2013,sharpnack2013b,sharpnack2015}, which certify that such procedures are \emph{optimal} over ``function'' classes $\mc{F} \subseteq \Reals^n$. The downside of this work is that it may be unnatural to assume a priori that the graph $G$ is fixed, or that the evaluations $\bigl(f_0(X_1),\ldots,f_0(X_n)\bigr)$ belong to some discrete ``function'' class.
	\item In the arguably more natural \emph{random design perspective},  one assumes the design points $X_1,\ldots,X_n$ are sampled independently from some distribution $P$ supported on a domain $\Xset \subset \Rd$. Inference is drawn on the regression function $f_0: \Xset \to \Reals$, which is typically assumed to be smooth in some continuum sense, e.g. it possesses a first derivative bounded in $L^{\infty}$ (H\"{o}lder) or $\Leb^2$ (Sobolev) norm. To conduct graph-based inference, the user first builds a neighborhood graph over the random design points---so that $W_{ij}$ is large iff $X_i$ and $X_j$ are close in say Euclidean distance---and then solves e.g.~\eqref{eqn:laplacian_smoothing} or~\eqref{eqn:laplacian_smoothing_test}. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a \emph{continuum} limit (see \citet{belkin07,trillos2018,vonluxburg2008} among others). However, until recently such statements were not accompanied by error rates, and such error rates as have been proved \citep{lee2016,trillos2020} are not optimal over continuum function spaces, such as H\"{o}lder or Sobolev classes. 
\end{itemize}
Hereafter, we will adopt the random design perspective, and seek to answer the following question:
\begin{quote}{When $f_0$ is assumed to smooth in a continuum sense, does Laplacian smoothing achieve optimal performance for estimation and goodness-of-fit testing?} 
\end{quote}
On the one hand, this is no small question---rather, it is arguably the fundamental question of nonparametric regression, and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to its competitors. On the other hand, it seems difficult to answer: there is a fundamental gap between the \emph{discrete} smoothness imposed by the penalty $f^T \Lap f$ in~\eqref{eqn:laplacian_smoothing} and the \emph{continuum} smoothness assumed on $f_0$, and to obtain sharp upper bounds we will need to bridge this gap in a tight sense.

@@

\paragraph{Advantages of the discrete approach.} In light of this last point, it is worth asking whether there is any \emph{statistical} advantage to solving a discrete problem such as~\eqref{eqn:laplacian_smoothing} (setting aside computational considerations for the moment). After all, we could have instead solved the following variational problem:
\begin{equation}
\label{eqn:thin_plate_spline}
\wt{f} := \argmin_{f} \biggl\{\sum_{i = 1}^{n} \bigl(Y_i - f(X_i)\bigr)^2 + \rho \cdot \int_{\Xset} \|\nabla f(x)\|^2 \,dx \biggr\},
\end{equation}
with the minimum computed over all functions $f$ for which the criterion is finite--or used the statistic
\begin{equation}
\label{eqn:thin_plate_spline_test}
\wt{T} := \|\wt{f}\|_n^2.
\end{equation}
The penalty in~\eqref{eqn:thin_plate_spline} leverages the assumption that $f_0$ possesses a bounded derivative in a seemingly natural manner. Indeed, for this reason the estimator $\wt{f}$ (and to a lesser extent, the statistic $\wt{T}$) are well known--- in the univariate case, $\wt{f}$ is simply a \emph{smoothing spline}, and when $d > 1$ it is an example of a \emph{thin-plate spline}---and their statistical properties are well-understood \citep{vandergeer2000, liu2019}. Moreover, there exist strong connections between smoothing/thin-plate splines and Laplacian smoothing: as we show later on in~\eqref{eqn:seminorm_consistency}, in a certain sense the Laplacian smoothing problem~\eqref{eqn:laplacian_smoothing} can be seen as a discrete and noisy approximation to~\eqref{eqn:thin_plate_spline}. At first blush, this suggests that Laplacian smoothing should at best inherit the statistical properties of smoothing/thin-plate splines, and at worst may have meaningfully larger error.

However, as we shall see the actual story is quite different: remarkably, Laplacian smoothing enjoys optimality properties in cases where the thin-plate spline problem~\eqref{eqn:thin_plate_spline} is not even well-posed. Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates} summarize the gap. As we establish in Theorems~\ref{thm:laplacian_smoothing_estimation1}-\ref{thm:laplacian_smoothing_testing_manifold}, when computed over an appropriately formed neighborhood graph, Laplacian smoothing estimators and tests are minimax optimal over first-order \emph{continuum} Sobolev balls. This holds true either when $\Xset \subset \Rd$ is a full-dimensional domain and $d = 1,2$ or $3$, or when $\Xset$ is a submanifold of $\Rd$ of intrinsic dimension $m = 1,2,$ or $3$. Additionally, the estimator $\wh{f}$ is nearly minimax optimal (to within a $(\log n)^{1/3}$ factor) when $d$ (or $m$) = $4$. By contrast, smoothing splines are optimal only when $d$ (or $m$) = $1$. In higher dimensions, technically speaking the problem~\eqref{eqn:thin_plate_spline} is not even well-posed, and any practical implementation of the estimator $\wt{f}$ will be inconsistent \citep{green93}. Hence, our results favorably distinguish Laplacian smoothing methods from their natural variational alternatives.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing} & Thin-plate splines \eqref{eqn:thin_plate_spline} \\
			\hline
			$d = 1$ & ${\bf n^{-2/3}}$ & \textcolor{red}{${\bf n^{-2/3}}$} \\
			$d = 2,3$ & ${\bf n^{-2/(2 + d)}}$ & \textcolor{red}{$1$} \\
			$d = 4$ & ${\bf n^{-1/3}} (\log n)^{1/3}$ & \textcolor{red}{$1$} \\
			$d \geq 5$  & $(\log n/n)^{4/(3d)}$ &\textcolor{red}{$1$} \\
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over first-order Sobolev balls. Rates in black are new results of this paper, and bolded rates are minimax optimal. Although we suppress it for conciseness, in all cases the dependence of the error rate on the radius of the Sobolev ball is also optimal. For thin-plate splines, ``\textcolor{red}{$1$}'' indicates that the estimator is inconsistent. When $\Xset$ is an $m$ dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$, without any change in the method in the case of Laplacian smoothing.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing_test} & Thin-plate splines \eqref{eqn:thin_plate_spline_test} \\
			\hline
			$d = 1$ & ${\bf n^{-4/5}}$ & \textcolor{red}{${\bf n^{-4/5}}$} \\
			$d = 2,3$ & ${\bf n^{-4/(4 + d)}}$ & \textcolor{red}{$n^{-1/2}$} \\
			$d \geq 4$\footnotemark & ${ \bf n^{-1/2}}$ & \textcolor{red}{${ \bf n^{-1/2}}$}
		\end{tabular}
	\end{center}
	\caption{Summary of testing rates over fixed radius first-order Sobolev balls; black, red, and bold font used as in Table~\ref{tbl:estimation_rates}. Rates for $d \geq 4$ assume $f_0 \in \Leb^4(\Xset,M)$. When $\Xset$ is an $m$ dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$.}
	\label{tbl:testing_rates}
\end{table}
\footnotetext{Assuming $f_0 \in \Leb^4(\Xset,M)$.}
\paragraph{Future extensions.}
To be clear, these results are not the end of the story. For one thing, the estimator $\wh{f}$ is defined only at the design points $\{X_1,\ldots,X_n\}$. As such, we measure loss using the in-sample mean squared error 
\begin{equation*}
\|\wh{f} - f_0\|_n^2 := \frac{1}{n}\sum_{i = 1}^{n} \Bigl(\wh{f}_i - f_0(X_i)\Bigr)^2.
\end{equation*}
In Section~\ref{sec:minimax_optimal_laplacian_smoothing}, we briefly discuss how one could extend $\wh{f}$ to a function over all $\Xset$, in such a way that the out-of-sample mean square error $\|\wh{f} - f_0\|_{\Leb^2(\Xset)}$ remains small, but leave a formal analysis of the problem to future work.

In a different direction, the problem~\eqref{eqn:thin_plate_spline} is only a special, first-order case of thin-plate splines. In general, the degree-$k$ thin plate spline is defined as a minimizer of
\begin{equation*}
\sum_{i = 1}^{n} (Y_i - f(X_i))^2 + \rho \cdot \sum_{|m| = k} \int_{\Xset} \bigl(D^mf(x)\bigr)^2 \,dx
\end{equation*}
where for each multiindex $m$ we write $D^m$ for the partial derivative $D^mf(x):= \partial^kf/\partial x_{1}^{m_1}\ldots \partial x_{d}^{m_d}$. Assuming the $k$th-order partial derivatives $D^mf_0$ are all bounded in $\Leb^2(\Xset)$ norm, the degree-$k$ thin plate spline has error on the order of $n^{-2k/(2k + d)}$ \citep{vandergeer2000}, which is the minimax optimal rate for such functions. Of course, assuming $f_0$ possesses $k$ bounded derivatives for some $2k > d$ is a rather strong assumption which may not be warranted, but at present we do not know whether (adaptations of) Laplacian smoothing on neighborhood graphs can achieve these faster rates.

% \paragraph{Notation.}
% AJG: Need to make sure these are all defined somewhere
% For an integer $p > 0$, we use $\Leb^p(\Xset)$ to refer to the set of functions $f$ for which the $p$th moment $\norm{f}_{\Leb^p(\Xset)}^p := \int_{\Xset} |f(x)|^p \,dx < \infty$, and $C^p(\Xset)$ to refer to those function which are $p$-times continuously differentiable with bounded $p$th derivative. Finally, we use $C,C_1,\ldots$ and $c,c_2,\ldots$ as constants which are independent of $n$ and $f_0$, and for convenience, we will also use asymptotic notation: $a_n \lesssim b_n$ means that $a_n \leq Cb_n$ for some $C > 0$, and $a_n \asymp b_n$ means that $a_n \lesssim b_n$ and $b_n \lesssim a_n$.

\section{BACKGROUND}
\label{sec:problem_setup_and_background}
Before we formally state our main results in Section~\ref{sec:minimax_optimal_laplacian_smoothing}, we define neighborhood graph Laplacians, and review the known minimax rates over first-order Sobolev spaces.

>>>>

\documentclass[twoside]{article}
\usepackage[accepted]{aistats2021}
\usepackage{amsmath, amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{verbatim}
\usepackage{microtype}
\usepackage{bm}

@@

\newcommand{\Linv}{L^{\dagger}}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\esssup}{ess\,sup}

@@

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}

@@

	\aistatsauthor{Alden Green \And Sivaraman Balakrishnan \And Ryan J. Tibshirani}

	\aistatsaddress{Carnegie Mellon University \And Carnegie Mellon University \And Carnegie Mellon University}]

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to nonparametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator \smash{$\wh{f}$}, and a goodness-of-fit test based on \smash{$\wh{f}$}. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, for $\Xset \subseteq \Rd$ and $1 \leq d < 4$; in the estimation problem, for $d = 4$, they are optimal modulo a $\log n$ factor. Additionally, we prove that Laplacian smoothing is manifold-adaptive: if $\Xset \subseteq \R^d$ is an $m$-dimensional manifold with $m < d$, then the error rate of Laplacian smoothing (in either estimation or testing) depends only on $m$, in the same way it would if $\Xset$ were a full-dimensional set in $\Reals^m$.
\end{abstract}

\section{INTRODUCTION}

We adopt the standard setup in nonparametric regression, where we observe samples $(X_1,Y_1),\ldots,(X_n,Y_n)$ that are i.i.d.\ draws from the model 
\begin{equation}
\label{eqn:signal_plus_noise_model}
Y_i = f_0(X_i) + \varepsilon_i, \quad \epsilon_i \sim N(0,1), 
\end{equation}
where $\varepsilon_i$ is independent of $X_i$. Our goal is to perform statistical inference on the unknown regression function $f_0$, by which we mean either \emph{estimating} $f_0$ or more simply \emph{testing} whether $f_0 = 0$, i.e., whether there is any signal present.

Laplacian smoothing \citep{smola2003} is a penalized least squares estimator, defined over a graph. Let $G = (V,W)$ be a weighted undirected graph with vertices $V=\{1,\ldots,n\}$, associated with $\{X_1,\ldots,X_n\}$, the Laplacian smoothing estimator \smash{$\wh{f}$} is given by
\begin{equation}
\label{eqn:laplacian_smoothing}
\wh{f} =  \argmin_{f \in \Reals^n} \; \sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^\top \Lap f. 
\end{equation}
Here $\Lap$ is the graph Laplacian matrix (to be defined in Section~\ref{sec:problem_setup_and_background}), $G$ is typically a geometric graph (such as a $k$-nearest-neighbor or neighborhood graph), $\rho \geq 0$ is a tuning parameter, and the penalty
\begin{equation*}
f^\top \Lap f = \frac{1}{2} \sum_{i,j = 1}^{n} W_{ij}(f_i - f_j)^2
\end{equation*}
encourages \smash{$\wh{f}_i \approx \wh{f}_j$} when $X_i \approx X_j$. Assuming \eqref{eqn:laplacian_smoothing} is a reasonable estimator of $f_0$, the statistic
\begin{equation}
\label{eqn:laplacian_smoothing_test}
\wh{T} = \frac{1}{n} \| \wh{f} \|_2^2 
\end{equation}
would in turn be a reasonable way to test $f_0 = 0$.

Of course there are many methods for nonparametric regression (see, e.g., \citet{gyorfi2006,wasserman2006,tsybakov2008_book}), but Laplacian smoothing has its own set of advantages. For instance:
\begin{itemize}
	\item \emph{Computational ease.} Laplacian smoothing is fast, easy, and stable to compute. The estimate \smash{$\wh{f}$} can be computed by solving a symmetric diagonally dominant linear system. There are by now various nearly-linear-time solvers for this problem (see e.g., the seminal papers of \citet{spielman2011,spielman2013,spielman2014}, or the overview by \citet{vishnoi2012} and references therein).
	\item \emph{Generality.} Laplacian smoothing is well-defined whenever one can associate a graph with observed responses. This generality lends itself to many different data modalities, e.g., text and image classification, as in \citet{kondor2002,belkin03a,belkin2006}.
	\item \emph{Weak supervision.} Although we study Laplacian smoothing in the supervised problem setting \eqref{eqn:signal_plus_noise_model}, the method can be adapted to the semi-supervised or unsupervised settings, as in \citet{zhu2003semisupervised,zhou2005learning,nadler09}. %,slepcev17,calder2019b,dunlop2020}.
\end{itemize}

For these reasons, a body of work has emerged that analyzes the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, this work can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item \emph{Fixed design perspective.} Here one treats the design points $X_1,\ldots,X_n$ and the graph $G$ as fixed, and carries out inference on $f_0(X_i)$, $i=1,\ldots,n$. In this problem setting, tight upper bounds have been derived on the error of various graph-based methods (e.g., \citet{wang2016,hutter2016,sadhanala16,sadhanala17,kirichenko2017,kirichenko2018}) and tests (e.g., \citet{sharpnack2010identifying,sharpnack2013b,sharpnack2013,sharpnack2015}), which certify that such procedures are \emph{optimal} over ``function'' classes (in parentheses because these classes really model the $n$-dimensional vector of evaluations). The upside of this work is its generality: in this setting $G$ need not be a geometric graph, but in principle it could be any graph over $V=\{1,\ldots,n\}$. The downside is that, in the context of nonparametric regression, it is as not natural to think of the evaluations of $f_0$ as exhibiting smoothness over some fixed pre-defined graph $G$, and more natural to speak of the smoothness of the function $f_0$ itself. 
	\item \emph{Random design perspective.} Here one treats the design points $X_1,\ldots,X_n$ as independent samples from some distribution $P$ supported on a domain $\Xset \subseteq \Rd$. Inference is drawn on the regression function $f_0: \Xset \to \Reals$, which is typically assumed to be smooth in some \emph{continuum} sense, e.g., it possesses a first derivative bounded in $L^{\infty}$ (H\"{o}lder) or $\Leb^2$ (Sobolev) norm. To conduct graph-based inference, the user first builds a neighborhood graph over the random design points---so that $W_{ij}$ is large when $X_i$ and $X_j$ are close in (say) Euclidean distance---and then computes e.g., \eqref{eqn:laplacian_smoothing} or \eqref{eqn:laplacian_smoothing_test}. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a continuum limit (see \citet{belkin07,vonluxburg2008,trillos2018} among others). However, until recently such statements were not accompanied by error rates, and even so, such error rates as have been proved \citep{lee2016,trillos2020} are not optimal over continuum function spaces, such as H\"{o}lder or Sobolev classes. 
\end{itemize}
The random design perspective bears a more natural connection with nonparametric regression (the focus in this paper), as it allows us to formulate smoothness based on $f_0$ itself (how it behaves as a continuum function, and not just its evaluations at the design points). In this paper, we will adopt the random design perspective, and seek to answer the following question:
\begin{quote}{When $f_0$ is assumed to smooth in a continuum sense, does Laplacian smoothing achieve optimal performance for estimation and goodness-of-fit testing?} 
\end{quote}
This is no small question---arguably, it is \emph{the} central question of nonparametric regression---and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to alternative methods. It also seems difficult to answer: as we discuss next, there is a fundamental gap between the \emph{discrete} smoothness imposed by the penalty $f^\top \Lap f$ in problem \eqref{eqn:laplacian_smoothing} and the \emph{continuum} smoothness assumed on $f_0$, and in order to obtain sharp upper bounds we will need to bridge this gap in a suitable sense.

@@

\paragraph{Advantages of the Discrete Approach.}

In light of this last point, it is worth asking whether there is any \emph{statistical} advantage to solving a discrete problem such as \eqref{eqn:laplacian_smoothing} (setting aside computational considerations for the moment). After all, we could have instead solved the following variational problem:
\begin{equation}
\label{eqn:thin_plate_spline}
\wt{f} = \argmin_{f : \Xset \to \Reals} \; \sum_{i = 1}^{n} \bigl(Y_i - f(X_i)\bigr)^2 + \rho \hspace{-2pt} \int_{\Xset} \|\nabla f(x)\|_2^2 \,dx,
\end{equation}
where the optimization is performed over all functions for which the criterion is well-defined and finite (continuous functions $f$ that have a weak derivative $\nabla f$ in $L^2(\Xset)$). Analogously, for testing, we could use:
\begin{equation}
\label{eqn:thin_plate_spline_test}
\wt{T} = \| \wt{f} \|_n^2 = \frac{1}{n} \sum_{i=1}^n \wt{f}(X_i)^2.
\end{equation}
The penalty term in \eqref{eqn:thin_plate_spline} leverages the assumption that $f_0$ has a smooth derivative in a seemingly natural way. Indeed, the estimator \smash{$\wt{f}$} and statistic \smash{$\wt{T}$} are well-known: for $d = 1$, \smash{$\wt{f}$} is the familiar \emph{smoothing spline}, and for $d>1$, it is type of a \emph{thin-plate spline}. The statistical properties of smoothing and thin-plate splines are well-understood \citep{vandergeer2000,liu2019}. As we discuss later, the Laplacian smoothing problem \eqref{eqn:laplacian_smoothing} can be viewed as a discrete and noisy approximation to \eqref{eqn:thin_plate_spline}. At first blush, this suggests that Laplacian smoothing should at best inherit the statistical properties of \eqref{eqn:thin_plate_spline}, and at worst may have meaningfully larger error.

However, as we shall see the actual story is quite different: remarkably, Laplacian smoothing enjoys optimality properties even in settings where the thin-plate spline estimator \eqref{eqn:thin_plate_spline} is not well-posed (to be explained shortly); Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates} summarize. As we establish in Theorems~\ref{thm:laplacian_smoothing_estimation1}-\ref{thm:laplacian_smoothing_testing_manifold}, when computed over an appropriately formed neighborhood graph, Laplacian smoothing estimators and tests are minimax optimal over first-order \emph{continuum} Sobolev balls. This holds true either when $\Xset \subseteq \Rd$ is a full-dimensional domain and $d = 1,2,$ or $3$, or when $\Xset$ is a manifold embedded in $\Rd$ of intrinsic dimension $m = 1,2,$ or $3$. Additionally, the estimator \smash{$\wh{f}$} is nearly minimax optimal (to within a \smash{$(\log n)^{1/3}$} factor) when $d=4$ (or $m=4$ in the manifold case).

By contrast, smoothing splines are optimal only when $d=1$. When $d>1$, thin-plate spline estimator \eqref{eqn:thin_plate_spline} is not even well-posed, in the following sense: for any $(X_1,Y_1),\ldots,(X_n,Y_n)$ and any $\delta>0$, there exists (e.g., \citet{green93} give a construction using ``bump'' functions) a differentiable function $f$ such that $f(X_i) = Y_i$, $i=1,\ldots,n$, and
\begin{equation*}
\int_{\Xset} \|\nabla f(x)\|_2^2 \leq \delta.
\end{equation*}
i.e., $f$ achieves perfect (zero) data loss and arbitrarily small penalty in the problem \eqref{eqn:thin_plate_spline}. This will clearly not lead to a consistent estimator of $f_0$ across the design points (as it always yields $Y_i$ at each $X_i$). In this light, our results for $d>1$ favorably distinguish Laplacian smoothing from its natural variational analog.

\begin{table}
	\begin{center}
		\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing} & Thin-plate splines \eqref{eqn:thin_plate_spline} \\
			\hline
			$d = 1$ & $\bm{n^{-2/3}}$ & \textcolor{red}{$\bm{n^{-2/3}}$} \\
			$d = 2,3$ & $\bm{n^{-2/(2 + d)}}$ & \textcolor{red}{$1$} \\
			$d = 4$ & $\bm{n^{-1/3}} (\log n)^{1/3}$ & \textcolor{red}{$1$} \\
			$d \geq 5$  & $(\log n/n)^{4/(3d)}$ &\textcolor{red}{$1$} \\
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over first-order Sobolev balls. Black font marks new results from this paper, red font marks previously-known results; bold font marks minimax optimal rates. Although we suppress it for simplicity, in all cases the dependence of the error rate on the radius of the Sobolev ball is also optimal. We use ``$1$'' to indicate inconsistency (error not converging to 0). Furthermore, when $\Xset$ is an $m$-dimensional manifold embedded in $\Rd$, all Laplacian smoothing results hold with $d$ replaced by $m$, without any change to the method itself.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing_test} & Thin-plate splines \eqref{eqn:thin_plate_spline_test} \\
			\hline
			$d = 1$ & $\bm{n^{-4/5}}$ & \textcolor{red}{$\bm{n^{-4/5}}$} \\
			$d = 2,3$ & $\bm{n^{-4/(4 + d)}}$ & \textcolor{red}{$n^{-1/2}$} \\
			$d \geq 4$ & $\bm{n^{-1/2}}$ & \textcolor{red}{$\bm{n^{-1/2}}$}
		\end{tabular}
	\end{center}
	\caption{Summary of testing rates over first-order Sobolev balls; black, red, and bold fonts are used as in Table~\ref{tbl:estimation_rates}. Rates for $d \geq 4$ assume that $f_0 \in \Leb^4(\Xset,M)$. When $\Xset$ is an $m$-dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$.}
	\label{tbl:testing_rates}
\end{table}
%\footnotetext{Assuming $f_0 \in \Leb^4(\Xset,M)$.}

\paragraph{Future Directions.}

To be clear, there is still much left to be investigated. For one, the Laplacian smoothing estimator \smash{$\wh{f}$} is only defined at $X_1,\ldots,X_n$. In this work we study its in-sample mean squared error 
\begin{equation}
\label{eqn:empirical_norm_error}
\bigl\|\wh{f} - f_0\bigr\|_n^2 := \frac{1}{n}\sum_{i = 1}^{n} \Bigl(\wh{f}_i - f_0(X_i)\Bigr)^2.
\end{equation}
In Section~\ref{sec:minimax_optimal_laplacian_smoothing}, we discuss how to extend \smash{$\wh{f}$} to a function over all $\Xset$, in such a way that the out-of-sample mean squared error \smash{$\|\wh{f} - f_0\|_{\Leb^2(\Xset)}^2$} should remain small, but leave a formal analysis to future work.

In a different direction, problem \eqref{eqn:thin_plate_spline} is only a special, first-order case of thin-plate splines. In general, the $k$th order thin-plate spline estimator is defined as
\begin{equation*}
\wt{f} = \argmin_{f : \Xset \to \Rd} \; \sum_{i = 1}^{n} \bigl(Y_i - f(X_i)\bigr)^2 + \rho \hspace{-2pt} \sum_{|\alpha| = k} \int_{\Xset} \bigl(D^\alpha f(x)\bigr)^2 \,dx,
\end{equation*}
where for each multi-index $\alpha=(\alpha_1,\ldots,\alpha_d)$ we write $D^\alpha f(x) = \partial^kf/\partial x_{1}^{\alpha_1} \cdots \partial x_{d}^{\alpha_d}$. This problem is in general well-posed whenever $2k > d$. In this regime, assuming that the $k$th order partial derivatives $D^\alpha f_0$ are all $\Leb^2(\Xset)$ bounded, the degree $k$ thin-plate spline has error on the order of \smash{$n^{-2k/(2k + d)}$} \citep{vandergeer2000}, which is minimax rate-optimal for such functions. Of course, assuming $f_0$ has $k$ bounded derivatives for some $2k > d$ is a very strong condition, but at present we do not know if (adaptations of) Laplacian smoothing on neighborhood graphs achieve these rates.

\paragraph{Notation.}

For an integer $p \geq 1$, we use $\Leb^p(\Xset)$ for the set of functions $f$ such that 
\begin{equation*}
\norm{f}_{\Leb^p(\Xset)}^p := \int_{\Xset} |f(x)|^p \,dx < \infty,
\end{equation*}
and $C^p(\Xset)$ for the set of functions that are $p$ times continuously differentiable. For sequences $a_n,b_n$, we write $a_n \lesssim b_n$ to mean $a_n \leq Cb_n$ for a constant $C>0$ and large enough $n$, and $a_n \asymp b_n$ to mean $a_n \lesssim b_n$ and $b_n \lesssim a_n$. Lastly, we use $a \wedge b = \max\{a,b\}$.

\section{BACKGROUND}
\label{sec:problem_setup_and_background}

Before we present our main results in Section~\ref{sec:minimax_optimal_laplacian_smoothing}, we define neighborhood graph Laplacians, and review known minimax rates over first-order Sobolev spaces.

\paragraph{Neighborhood Graph Laplacians.}
